Asymmetric tail association in extreme climatic events affects population-extinction risk

### 2019-10-14 MainText.Rmd
methods editing continued

---
title: "Asymmetric tail-association in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs 
could pose severe impact on the resistance or resilience of a population and/ or community level through
life-history traits of the species. Therefore predicting the impact of extreme events 
on population extinction risk is necessery. We here introduce the potential use of "tail-association" 
property of two correlated ecological variables in modeling the extinction risk of 
an age-structured single patch model. We also extend this idea 
to density-dependent metapopulation models. For metapopulation models, 
we show extinction risk not only depends on the nature of tail-association of 
spatially correlated noise but is also influenced by the inherent dynamics 
of the population-model. To illaborate, we demonstrate left tail dependent noise across spatially correlated patches 
cause higher extinction risk than the right tail dependent noise, when the model show undercompensatory dynamics.
But in overcompensatory regime, the trend is opposite. Our study paves the path for future research considering this tail-association instead of simple correlation in population viability analysis.

@@

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated variables are more associated in their larger values than smaller, they are 
said to have stronger "right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows two random variables can have a wide 
range of assciation from extreme to moderate degree in one of their 
extreme tails - often refers to having 'asymmetric tail' - but with a 
fixed Kendall's $\tau$. One can control the degree of "tail-asymmetry" - moderate 
to extreme - using "copula", a statistical 
tool generally used to model dependencies between correlated variables (for details see @Ghosh_copula and \nameref{Methods}). 
This 'asymmetric tail' is commonly found in many 
bivariate or multivariate dependence structures for environmental, species-trait, 
phenological, population, community, and ecosystem functioning datasets 
and can help to interpret the causal mechanisms as well as 
to draw consequential inferences for such patterns [@Ghosh_copula]. For example, it 
is well known that spatial dependence of environmental
variables is an important driver of spatial dependence in fluctuating 
population time series (called 'spatial synchrony'). Therefore, one may expect 
the synchronous populations over a larger network would suppress for extreme bad 
years of limiting resources than moderately 'good' years. Any two such populations, that are 
affected badly in a similar manner due to common food shortage, are expected to be more
"left-tail" associated (Figure \ref{fig_intro_cop}A, B) rather than "right-tail" (Figure \ref{fig_intro_cop}C, D) as the environmental fluctuation 
time-series have more "left-tail" association for those extreme bad years
(i.e., 'Moran-effect' like mechanism). Similarly, during 
favorable years if pest metapopulations 
grow abundantly compared to other average years, population time series look like 
\ref{fig_intro_cop}C or D with stronger "right-tail" association, turning local 
outbreaks into regional epidemics. In a 
nutshell, "tail-asymmetry" could have profound effect in understanding 
risk for population viability or ecosystem function
considering their relation with extreme climatic events.

A recent study [@Ghosh_copula] used the above mentioned feature of "tail-asymmetry" 
to estimate extinction risk for a density-independent metapopulation model. The study 
showed "left-tail" associated density-independent metapopulations 
to be more prone to extinctions than "right-tail" associated one as we may expect species 
are more vulnerable to "catastrophs" (extreme bad years) than to "bonanzas" (extreme good years).
Present manuscript is an extension of that concept in estimating the extinction risk 
beyond the density-independent model. Population growth rate, when depending 
on its own density, serves as a more realistic model to regulate unbounded growth, hence 
models with "density-dependence" are often studied in population ecology to 
predict population viability over long-term period [@hassell1975density; @fowler1987review; 
@courchamp1999inverse]. Here, we will explore the general finding from @Ghosh_copula in the context of
the density-dependent metapopulation patches which experience environmental noises with same spatial correlation but with a variety of tail 
asymmetry (from extreme to moderate, Fig. \ref{fig_intro_cop}). It is expected 
that extreme left-tail associated noise might wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

Tail-asymmetry in estimating extinction risk can also be applied to an age structured model. From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate (fecundity) as well as larval survival. This scenario would resemble Fig.\ref{fig_intro_cop}(A, B) with stronger tail association between vital rates of two stages at their smaller values than larger ones and thereby inflate the population viability compared to a scenario as of Fig.\ref{fig_intro_cop}(C, D). <!--***Shyamolina: Should I tell this? 
It is reasonable to assume cold winter-temperature as the limiting factor 
for each age-structured population - thereby also showing "left-tail" asymmetry between 
the temperature and each age-population time series.(Ref: driver-synchrony relationship as shown in our aphid-plankton paper)--> Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs with age structured 
model or density-dependent metapopulations.

@@

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-association 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail association 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail association across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail association) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how does asymmetric tail 
dependent noise affect the extinction risk of
populations in a single patch age-structured model,
we considered a stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear".
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail-association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates. Vital rates ($f_a$ and $s_e$) were
left tail dependent when unusual bad years ("catastrophs") suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

@@

\begin{equation}
f_a = \Gamma (k, \theta)\exp(-A/\Omega), \hspace{0.2 cm} s_e = \mathcal{B}(\alpha,\beta) \label{eq_vitalrates}
\end{equation}

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail association in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail association
with positive Kendall's correlation ($\tau = 0.6$).

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we feed pairwise entries from each row of matrix $M$ as the vital rates 
(fecundity $f_a$ and survival rates $s_e$) 
of the stochastic projrction matrix in model (\ref{eq_age_str}).

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail association of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-association. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with ($t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and extreme (or moderate) left 
tail- and right-tail dependent environmental noise 
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-association (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail-association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

<!--summary results:-->
We showed the potential use of "asymmetric" tail association in modeling the impact of 
extreme ecological events on population-extinction risk. Our results demonstrated left and right-tail 
association in vital rates (or in spatially correlated environmental noises) affected extinction risk differently
in years of extreme events ("catastrophs" and "bonanzas"). Though we could expect left tail association 
to show higher extinction risk than the one for right tail association with age-structured 
and density dependent metapopulation models, it was not always true. In the over-compensatory regime,
right tail dependent noise caused higher extinction risk than the left tail dependent noise. We showed 
that in absence of dispersal the above-said switching took place near the critical parameter value where 
the deterministic model showed transition from under-compensatory to over-compensatory regime.

>>>>

---
title: "Asymmetric tail association in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs 
could pose severe impact on the resistance or resilience of a population and/ or community level through
life-history traits of the species. Therefore predicting the impact of extreme events 
on population extinction risk is necessery. We here introduce the potential use of "tail association" 
property of two correlated ecological variables in modeling the extinction risk of 
an age-structured single patch model. We also extend this idea 
to density-dependent metapopulation models. For metapopulation models, 
we show extinction risk not only depends on the nature of tail association of 
spatially correlated noise but is also influenced by the inherent dynamics 
of the population-model. To illaborate, we demonstrate left tail dependent noise across spatially correlated patches 
cause higher extinction risk than the right tail dependent noise, when the model show undercompensatory dynamics.
But in overcompensatory regime, the trend is opposite. Our study paves the path for future research considering this tail association instead of simple correlation in population viability analysis.

@@

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated variables are more associated in their larger values than smaller, they are 
said to have stronger "right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows two random variables can have a wide 
range of assciation from extreme to moderate degree in one of their 
extreme tails - often refers to having 'asymmetric tail' - but with a 
fixed Kendall's $\tau$. One can control the degree of "tail asymmetry" - moderate 
to extreme - using "copula", a statistical 
tool generally used to model dependencies between correlated variables (for details see @Ghosh_copula and \nameref{Methods}). 
This 'asymmetric tail' is commonly found in many 
bivariate or multivariate dependence structures for environmental, species-trait, 
phenological, population, community, and ecosystem functioning datasets 
and can help to interpret the causal mechanisms as well as 
to draw consequential inferences for such patterns [@Ghosh_copula]. For example, it 
is well known that spatial dependence of environmental
variables is an important driver of spatial dependence in fluctuating 
population time series (called 'spatial synchrony'). Therefore, one may expect 
the synchronous populations over a larger network would suppress for extreme bad 
years of limiting resources than moderately 'good' years. Any two such populations, that are 
affected badly in a similar manner due to common food shortage, are expected to be more
"left-tail" associated (Figure \ref{fig_intro_cop}A, B) rather than "right-tail" (Figure \ref{fig_intro_cop}C, D) as the environmental fluctuation 
time-series have more "left-tail" association for those extreme bad years
(i.e., 'Moran-effect' like mechanism). Similarly, during 
favorable years if pest metapopulations 
grow abundantly compared to other average years, population time series look like 
\ref{fig_intro_cop}C or D with stronger "right-tail" association, turning local 
outbreaks into regional epidemics. In a 
nutshell, "tail asymmetry" could have profound effect in understanding 
risk for population viability or ecosystem function
considering their relation with extreme climatic events.

A recent study [@Ghosh_copula] used the above mentioned feature of "tail asymmetry" 
to estimate extinction risk for a density-independent metapopulation model. The study 
showed "left-tail" associated density-independent metapopulations 
to be more prone to extinctions than "right-tail" associated one as we may expect species 
are more vulnerable to "catastrophes" (extreme bad years) than to "bonanzas" (extreme good years).
Present manuscript is an extension of that concept in estimating the extinction risk 
beyond the density-independent model. Population growth rate, when depending 
on its own density, serves as a more realistic model to regulate unbounded growth, hence 
models with "density-dependence" are often studied in population ecology to 
predict population viability over long-term period [@hassell1975density; @fowler1987review; 
@courchamp1999inverse]. Here, we will explore the general finding from @Ghosh_copula in the context of
the density-dependent metapopulation patches which experience environmental noises with same spatial correlation but with a variety of tail 
asymmetry (from extreme to moderate, Fig. \ref{fig_intro_cop}). It is expected 
that extreme left-tail associated noise might wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

Tail asymmetry in estimating extinction risk can also be applied to an age structured model. From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate (fecundity) as well as larval survival. This scenario would resemble Fig.\ref{fig_intro_cop}(A, B) with stronger tail association between vital rates of two stages at their smaller values than larger ones and thereby inflate the population viability compared to a scenario as of Fig.\ref{fig_intro_cop}(C, D). <!--***Shyamolina: Should I tell this? 
It is reasonable to assume cold winter-temperature as the limiting factor 
for each age-structured population - thereby also showing "left-tail" asymmetry between 
the temperature and each age-population time series.(Ref: driver-synchrony relationship as shown in our aphid-plankton paper)--> Though tail association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs with age structured 
model or density-dependent metapopulations.

@@

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail association 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail association 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How do asymmetric tail associations in environmental noises across multiple patches influence extinction risk of a density dependent metapopulation model, and how do the effects 
interact with the nature of density dependence? Will they remain consistent with the 
general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail associated than the one 
if noises had right-tail association) irrespective of the nature of density dependence?

<!--
How does environmental noises with asymmetric tail association across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail association) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.-->

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how do vital rates with asymmetric tail 
association affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$).

@@

\begin{equation}
f_a \sim \Gamma (k, \theta)\exp(-A/\Omega), \hspace{0.2 cm} s_e \sim \mathcal{B}(\alpha,\beta) \label{eq_vitalrates}
\end{equation}

We assumed the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear" with density dependence.
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates (as depicted in Fig. \ref{fig_intro_cop}). Vital 
rates ($f_a$ and $s_e$) were left tail dependent when unusual bad years ("catastrophes") 
suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

Moderate tail associatation was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail asscociations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copula had the same Kendall 
correlation ($\tau =$ `r ktau_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau =$ `r ktau_common`). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail association
with positive Kendall's correlation.

On a similar note, we generated extreme left- (or right-) tail dependent matrix (M)
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula. We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we incorporated pairwise entries from each row of matrix $M$ in the expression for 
vital rates (fecundity $f_a$ and survival rates $s_e$, Eq. \ref{eq_vitalrates}) of 
the stochastic projrction matrix in model (\ref{eq_age_str}).

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail association of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependent feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form: $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations 
between patches (to within sampling error)
for every simulation, but were moderately (or extremely) left- and 
right-tail associated between patches in
different simulations. For moderately left-tail associated 
noises we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($\tau =$ `r ktau_common`) between
any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail associated noises we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail association. We generated
extreme left- or right-tail associated noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$ 
that covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
to get a positive, finite equilibrium $P_e$ (for details, see Appendix 
\ref{SM-model_stability}) to start with (at $t=0$). For each chosen $r$,
we calculated extinction risk after $25$ years with each metapopulation model 
where population could disperse locally (or globally) with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$ and moderate to extreme left 
tail- and right-tail associated environmental noises with same Kendall's $\tau$ (= `r ktau_common`)
in different simulations (for code see [`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R)).

<!--start from here-->

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail association (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

<!--summary results:-->
We showed the potential use of "asymmetric" tail association in modeling the impact of 
extreme ecological events on population-extinction risk. Our results demonstrated left and right-tail 
association in vital rates (or in spatially correlated environmental noises) affected extinction risk differently
in years of extreme events ("catastrophes" and "bonanzas"). Though we could expect left tail association 
to show higher extinction risk than the one for right tail association with age-structured 
and density dependent metapopulation models, it was not always true. In the over-compensatory regime,
right tail dependent noise caused higher extinction risk than the left tail dependent noise. We showed 
that in absence of dispersal the above-said switching took place near the critical parameter value where 
the deterministic model showed transition from under-compensatory to over-compensatory regime.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_mtd}) 
unless the tail association was at its extreme (Fig. \ref{fig_extrisk_agestr_etd}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being dependent on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity - which was based
on the linear covariance measure between any two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, population growth rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail association.
